# Key Programming Concepts

## Lecture summary

This chapter builds a foundation for writing clear, reliable R code you can reuse across labs and projects. In lab we will practice these skills and discuss the weekly reading, Alston et al. 2020. A Beginner's Guide to Conducting Reproducible Research.

###  Learning objectives

By the end of this chapter, you will be able to:

- Distinguish **objects**, **functions**, and **expressions** in R, and use assignment idioms confidently.
- Work with core **data structures** (vectors, matrices/arrays, lists, and data frames) and explain when to use each.
- Use **control flow** (`if`, `for`) and the **apply** family to write clear, vectorized code.
- Write and call **functions** with arguments and return values; document intent with informative names.
- Perform basic **Boolean logic** and use logical indexing for subsetting.
- Read R help, search effectively, and follow reproducible coding practices you’ll build on all term.

###  Why programming concepts matter

Across languages, programming is built from a small set of ideas that you will reuse constantly. Greg Wilson (Software Carpentry) popularized a helpful list of seven: 

1.  Individual things (the number 2, the character 'hello')
2.  Commands that operate on things (the + symbol, the `length` function)
3.  Groups of things (R vectors, matrices, lists, dataframes, and arrays)
4.  Ways to repeat yourself (for and while loops, apply functions)
5.  Ways to make choices (if and try statements)
6.  Ways to create chunks (functions, objects/classes, and packages)
7.  Ways to combine chunks (piping) 

In this course we focus on doing these in **R** for ecological and conservation data.

###  Objects and basic types

In R, everything you create is an **object** that lives in your Global Environment. Objects have **classes** (e.g., numeric, integer, character), and functions behave differently depending on the class. Numbers come in two common flavors: `integer` and `numeric`. Numerics have very high—but finite—precision (about 15–16 decimal places). Text is stored as **character** strings, e.g., `"hello world"`.


###  Assignments vs. expressions

An **expression** *is* a thing; a **statement** *does* a thing. In R, you bind a name to a value with the assignment arrow: `hours_per_wk <- 168`. Read that as “make `hours_per_wk` refer to 168,” not as a math equation.


###  Functions and operators

A **function** is a reusable action you call by name, like `max(15, 27)`. Functions take **arguments** and often **return** a value. Many mathematical symbols are just functions in disguise (operators), e.g., `x + y` calls the `+` function. You can and should write your own functions to package work you do often.


###  Core data structures

You will manipulate collections of values constantly:

- **Vector**: a one-dimensional sequence where every element has the same class. Arithmetic is *vectorized*—it applies element-by-element.
- **Matrix**: a 2-D array (rows × columns) of a single class.
- **Array**: the N-D generalization of a matrix.
- **Data frame**: a table whose columns can be different classes but must all be the same length. Each row is an observation; each column is a variable.
- **List**: an ordered bag of components that can be different sizes and classes. A data frame is a special kind of list where all components are equal length.

###  Comparisons and boolean logic

Comparisons produce **logical** values: `TRUE` or `FALSE`. Operators include `>`, `>=`, `==` (equal), and `!=` (not equal); combine them with `&` (and) or `|` (or). Logical vectors are powerful: `sum(c(TRUE, FALSE, TRUE))` returns 2 because `TRUE` is 1 and `FALSE` is 0. 


###  Control flow: `if` and `for`

Control statements direct the path of your code. Use `if (...) { ... } else { ... }` to make choices, and `for (name in collection) { ... }` to repeat work for each element. Indentation isn’t required by R, but using it makes your intent clear. 


###  The *apply* family (vectorized repetition)

Many tasks that could be written with `for` loops are more concise and often faster with the *apply* family: `apply()` (arrays/matrices), `lapply()` (returns a list), `sapply()` (simplifies when possible), `mapply()` (multiple inputs), and `mclapply()` (parallel). Each takes a `FUN` argument—the function to apply to each element. 


###  Writing your own functions

Encapsulate a task by defining a function:

```r
spread <- function(values) {
  out <- max(values) - min(values)
  return(out)
}
```

Well-named functions make analysis pipelines clearer, easier to test, and reusable. 


###  Combining steps with pipes

Pipes let you express “do X, then Y, then Z” by chaining operations. We will pipe `%>%` from the tidyverse. You’ll see more next week when we start building data-wrangling workflows. 


###  R vs. Python: a quick orientation

R emerged from statistics; Python from computer science. Both are open-source and widely used in data work. We’ll use R throughout this course, but the concepts above transfer between languages. 

An annoying difference Python users will struggle with at first: in R, the syntax for calling a **function** and a **method** looks the same (because most things are functions), whereas Python distinguishes `data.plot` from `plot(data)`. 

```{r child="demos/01-demo-fundamentals.Rmd", echo=FALSE, error=TRUE}
```

```{r child="demos/02-demo-exponential-growth.Rmd", echo=FALSE, error=TRUE}
```

```{r child="readings/02-reading.Rmd", echo=FALSE, error=TRUE}
```

```{r child="labs/02-lab_populations_student.Rmd", echo=FALSE, error=TRUE}
```