# Using "The Tidyverse" for data manipulation and visualization

## Lecture summary

This chapter introduces the Tidyverse as a coherent grammar for data—wrangling with dplyr (`filter()`, `select()`, `mutate()`, `summarise()`, `group_by()`), reshaping with tidyr (`pivot_longer()`, `pivot_wider()`), joining relational tables, and visualizing with `ggplot2`. We build readable pipelines with the pipe, emphasize tidy data principles, and handle factors, dates, and missing values thoughtfully. In lab, we apply these tools to climate indicators (Mauna Loa CO₂, global temperature, sea ice extent, Antarctic mass), cleaning and merging heterogeneous files, pivoting wide↔long, computing windowed averages, and producing plots that separate trend, seasonality, and noise. We will read a best practice manual for writing code you and other people can read easily. 

### Learning Objectives

By the end of this chapter, you will be able to:

- Describe the key components of the Tidyverse and how they integrate for data analysis.
- Apply `dplyr` verbs for wrangling: `filter()`, `select()`, `mutate()`, `summarise()`, and `group_by()`.
- Reshape data between wide and long formats using `tidyr` (`pivot_longer()`, `pivot_wider()`).
- Create and customize visualizations with `ggplot2` (geoms, aesthetics, faceting, scales, themes).
- Combine datasets using joins in `dplyr` and choose appropriate join types for analysis goals.
- Interpret tidy summaries and plots to inform exploratory data analysis.


```{r co2_shared_setup, include=FALSE}
# Shared CO2 data for demos (uses the same file and parsing as the lab)
if (!exists("data_dir")) data_dir <- "data"
if (!exists("path_data")) path_data <- function(...) file.path(data_dir, ...)

co2_path <- path_data("co2_mm_mlo.txt")
has_co2 <- file.exists(co2_path)

if (has_co2) {
  # Read like the lab does (comment header + sentinel -9.99 -> NA)
  co2_raw <- read.table(co2_path, header = FALSE, comment.char = "#")
  names(co2_raw) <- c("year", "month", "decimal_date", "average", "interpolated",
                      "trend", "stdev_days", "uncertainty_average")
  co2_raw$average[co2_raw$average < 0] <- NA
  co2_raw$interpolated[co2_raw$interpolated < 0] <- NA
  co2_raw$date <- as.Date(sprintf("%d-%02d-01", co2_raw$year, co2_raw$month))
  if (requireNamespace("tibble", quietly = TRUE)) {
    co2_tbl <- tibble::as_tibble(co2_raw)
  } else {
    co2_tbl <- co2_raw
  }
} else {
  message("Demo 1 note: ", co2_path, " is missing. Place lab data in a 'data/' folder next to this Rmd.")
}

has_tidyverse <- requireNamespace("tidyverse", quietly = TRUE)
if (has_tidyverse) library(tidyverse)
knitr::opts_chunk$set(eval = has_tidyverse)
```

> If chunks don't run, install the meta-package: `install.packages("tidyverse")`

## Demo: Introduction to the Tidyverse

The **tidyverse** is a family of packages for modern data science in R. Loading the meta‑package attaches the most common tools (ggplot2, dplyr, tidyr, readr, tibble, stringr, forcats, purrr).

### Data wrangling

Core verbs from **dplyr** operate on tibbles and return new tibbles.
*What we’re doing.* Below we practice core dplyr verbs on the **CO₂** monthly series:

- `select()`/**`rename()`** to keep/rename a minimal set of columns, rename some
- `filter()`/`arrange()` to restrict to years post‑2000 months and find the highest 5 values
- `mutate()` to add derived features we’ll use later.

```{r}
# Select columns we are interested in and rename average for clarity 
co2_sub <- co2_tbl %>%
  dplyr::select(date, year, month, average) %>%
  dplyr::rename(avg_ppm = average)
```

```{r}
# Filter to 200+ with valid values; list the 5 highest months
co2_top <- co2_sub %>%
  dplyr::filter(year >= 2000, !is.na(avg_ppm)) %>%
  dplyr::arrange(dplyr::desc(avg_ppm)) %>%
  dplyr::slice_head(n = 5)
co2_top
```


```{r}
# Add derived columns: month name, monthly change, and decade
# R has a built in constant month.abb vector of the abbreviated month names.

co2_enriched <- co2_sub %>%
  dplyr::mutate(
    month_name = month.abb[month],
    delta = avg_ppm - dplyr::lag(avg_ppm),
    decade = paste0(floor(year/10)*10, "s")
  ) %>%
  dplyr::relocate(decade, .before = year)

head(co2_enriched)
```

#### Reshaping Data with tidyr — From Wide to Long and Back

Go from wide → long to make a tidy “series” column
Why: Most plotting and grouped summaries are easier when all measurements live in a single value column with a label telling you which measurement it is.
```{r}
co2_long <- co2_tbl %>%
  select(date, year, month, average, interpolated, trend) %>%
  pivot_longer(
    c(average, interpolated, trend),
    names_to  = "series",
    values_to = "ppmv"
  )

head(co2_long)

```
What this did: Stacked average, interpolated, and trend into one column ppmv, with a companion label series telling you which it was on each row.

Next, summarize by year, then long → wide for easy comparison
Why: Sometimes you want one column per series so you can compare or compute differences directly (e.g., average - trend).

```{r}
co2_yearly_wide <- co2_long %>%
  filter(!is.na(ppmv)) %>%
  group_by(year, series) %>%
  summarize(mean_ppm = mean(ppmv), .groups = "drop") %>%
  pivot_wider(
    names_from  = series,      # columns: average, interpolated, trend
    values_from = mean_ppm
  ) %>%
  arrange(year)

head(co2_yearly_wide)

```
What this did: Calculated annual means for each series, then spread them into separate columns so each row is a year with average, interpolated, and trend side-by-side.

Build compact keys with unite(), then recover fields with separate()
Why: It’s common to create an ID like "YYYY-MM" for joins or labeling, then later split it back into usable numeric pieces.

```{r}
# Make a compact "year-month" key
co2_ym <- co2_tbl %>%
  transmute(year, month, avg_ppm = average) %>%
  unite("ym", year, month, sep = "-", remove = FALSE)

head(co2_ym)

```
What this did: Created a ym key like "1988-07" while keeping year/month (because remove = FALSE).

Split the key back out: 
```{r}
# convert=TRUE turns strings into numbers
co2_ym_split <- co2_ym %>%
  separate(ym, into = c("year2", "month2"), sep = "-", convert = TRUE)

head(co2_ym_split)

```
What this did: Recovered year2 / month2 as numeric columns (handy if the key came from somewhere else and you need the parts back).

### Data visualization

*What we’re doing.* We’ll visualize the CO₂ series in two ways:
- a **line plot** of monthly CO₂ over time, and
- a **histogram of monthly change** (`delta`) faceted by decade to compare distributions.

```{r}
# Line: monthly CO₂ concentration
co2_enriched %>%
  ggplot2::ggplot(ggplot2::aes(date, avg_ppm)) +
  ggplot2::geom_line() +
  ggplot2::labs(x = "Date", y = "CO2 (ppm)", title = "Monthly CO2 at Mauna Loa") +
  ggplot2::theme_minimal()
```

```{r}
# Histogram: month-to-month change, by decade
co2_enriched %>%
  dplyr::filter(!is.na(delta)) %>%
  ggplot2::ggplot(ggplot2::aes(delta)) +
  ggplot2::geom_histogram(bins = 10) +
  ggplot2::facet_wrap(~ decade) +
  ggplot2::labs(x = "Δ ppm (current − previous month)", y = "Months",
                title = "Distribution of monthly CO2 changes by decade") +
  ggplot2::theme_minimal()

# The x-axes get cut off, which is annoying, but as an exploratory analysis we can leave it. 
```
What is your interpretation of these historgrams?

### Common types of visualizations

- Trends: `geom_line()`, `geom_smooth()`
- Distribution: `geom_histogram()`, `geom_density()`, `geom_boxplot()`, `geom_violin()`
- Relationship: `geom_point()` (+ `geom_smooth()`)
- Composition: stacked/filled bars (`geom_bar(position = "fill")`), area charts
- Uncertainty: `geom_errorbar()`, `geom_ribbon()`

### Fun graph references:
[Friends don't let friends make bad graphs](https://github.com/cxli233/FriendsDontLetFriends?tab=readme-ov-file)

[R graph gallary](https://r-graph-gallery.com/)

[R colors](https://www.nceas.ucsb.edu/sites/default/files/2020-04/colorPaletteCheatsheet.pdf)

---

```{r child="readings/03-reading.Rmd", echo=FALSE, error=TRUE}
```

---

```{r child="labs/03-lab-climate_student.Rmd", echo=FALSE, error=TRUE}
```
