## Demo: Create a graph from OBA data, visualize it

Let's load in our OBA data and make a network for all the bumble bees in Oregon. We will need to subset the data, then keep only the bee and plant species columns. 

```{r, load-oba}
oba <- read.csv("../data/OBA_2018-2024.csv")
glimpse(oba)
```

The entire network would be difficult to visualize, so let's subset the data to just bumble bees. We need to extract only the columns that are relevant for the bee-plant interactions. The column "speciesPlant" has the floral species a bee was caught visiting, if it was caught on a flower.

```{r, subset-obs}
## subset to only bombus
bumbles <- oba[oba$genus == "Bombus", ]

## drop rows without species names
bumbles <- bumbles[bumbles$specificEpithet != "",]

## create a bee genus species column 
bumbles$GenusSpecies <-  paste(bumbles$genus, bumbles$specificEpithet)

## subset down to just the columns of interest for a network
bumbles <- bumbles[, c("GenusSpecies", "speciesPlant")]

## rename them to something more consistent
colnames(bumbles) <- c("GenusSpecies", "PlantGenusSpecies")
head(bumbles)
```

There are a lot of blank rows from associated taxa for individuals not caught on a flower. We cannot do anything with those so we will just drop them. 

```{r, subset-obs-2}
bumbles <- bumbles[bumbles$PlantGenusSpecies != "",]
sort(unique(bumbles$PlantGenusSpecies))
sort(unique(bumbles$GenusSpecies))
```
You all are a very lucky class because it previous years there were a ton of terrible looking names.

We can explore the species in our network. These will be the vertices. 
```{r, explore}
sort(table(bumbles$GenusSpecies))
sort(table(bumbles$PlantGenusSpecies))
```
We have a variety of options for converting of data (which is basically an edge list) into a graph. One is to sum up our interactions but bee-plant combinations to make an bipartite adjacency matrix. Then convert that matrix to a igraph object. 

```{r, make-adj-1}
### 1. sum up interactions and take a look
bumbles_adj <- table(bumbles)
bumbles_adj <- unclass(bumbles_adj)  
### write to a file
save(bumbles_adj, file="bumbles_adj_mat.Rdata")
dim(bumbles_adj)
bumbles_adj[1:5, 1:5] 
```

```{r, make-adj-2}
### 2. convert to igraph bipartite adjacency matrix 
g_bumbles <- graph_from_biadjacency_matrix(bumbles_adj,
                                           weighted =TRUE) 
g_bumbles
```

To use the 'forceNetwork' function to make a beautiful visualization, we need to convert our graph into a dataframe. Luckily there is a function 'igraph_to_networkD3' to convert from graph objects to what the network3d package wants. That function also wants a each vertex assigned to a group. We have a variety of options on how to assign groups, one is to look for modules or compartments in the network. 

```{r, plot-oba}
### find modules
mod_bumbles <- cluster_walktrap(g_bumbles)
### assign vertices to modules
groups <- membership(mod_bumbles)

### convert to a network 3d object
g_bumbles_net3d <- igraph_to_networkD3(g_bumbles, group=groups, 
                                       what = "both")
### plot the bumble-plant network
forceNetwork(Links = g_bumbles_net3d$links,
             Nodes = g_bumbles_net3d$nodes,
            Source = "source", Target = "target",
            Value = "value",  NodeID = "name",
             Group = "group", 
            opacity = 0.8, zoom=TRUE, opacityNoHover = 0.9)
```

We can also use a flow-like visualization plot. 

```{r, plot-oba-2}

sankeyNetwork(Links = g_bumbles_net3d$links,
             Nodes = g_bumbles_net3d$nodes,
            Source = "source", Target = "target",
            Value = "value",  NodeID = "name", 
            nodePadding = 0, height = 5000)
```

### In class: Visualization challange 1.

Choose a different genus of bees and repeat the steps above to make a network. 

### In class: Visualization challange 2. 
How could we use graph attributes to create informative network visualizations? Set some attributes and use them to plot your network.
Consider using static network plots instead of network3D, see this [turorial](https://yunranchen.github.io/intro-net-r/advanced-network-visualization.html#visualization-for-static-network) for examples.  


```{r, plot-oba-challange}

```
